What's in a Name? \\ Linear Temporal Logic \\ Literally Represents Time Lines

### 2023-08-26 paper/main.tex
acknowledgements
suppress page number

\newcommand{\tool}{\hspace{0.1cm}\texttt{ltl2timeline}\hspace{0.1cm}}

@@

\title{What's in a Name? Linear Temporal Logic Literally Represents Time Lines}

\author{\IEEEauthorblockN{Runming Li\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
runmingl@andrew.cmu.edu}
\and
\IEEEauthorblockN{Keerthana Gurushankar\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
kgurusha@andrew.cmu.edu}
\and
\IEEEauthorblockN{Marijn~J.H.~Heule}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, USA \\
marijn@cmu.edu}
\and
\IEEEauthorblockN{Kristin~Y.~Rozier}
\IEEEauthorblockA{\textit{Iowa State University} \\
Ames, USA \\
ekyrozier@iastate.edu}
}

@@

\thispagestyle{plain}
\pagestyle{plain}

@@

\begin{definition}[Linear Temporal Logic (\ltl)]
    % cite Kristin's survey paper
    The syntax of an \ltl\ formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, is given by the following grammar:
    \begin{align*}
        \Phi ::= & p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi \\ & \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi
    \end{align*}\label{ltl-defn}
\end{definition}

\begin{remark}
    Intuitively, $\always \Phi$ says that formula $\Phi$ is true at every time step; $\eventually \Phi$ says that formula $\Phi$ is true either now or at sometime in the future time step; and $\nextt \Phi$ says formula $\Phi$ is true at the next time step immediately after the current one; $\Phi \stronguntil \Psi$ says that formula $\Phi$ is true \textit{until} such time formula $\Psi$ becomes true; $\Phi \weakrelease \Psi$ says that formula $\Psi$ must be true now and remain true until formula $\Phi$ becomes true, after which point \textit{release} $\Psi$ (if formula $\Phi$ is never true, then formula $\Psi$ must remain true at all time).
\end{remark}

@@

\begin{definition}[Semantics of \ltl\ formula]
    The semantics of \ltl\ formula is defined via  $\pi, i \models \Phi$ where $\pi : \omega \rightarrow 2^{\AP}$ is the computation that stores the truthhood and falsehood of every atomic proposition at every time step, and $\omega$ is the set of natural numbers that denote the time step. The semantics follows exactly from Rozier~\cite{Roz11}.
\end{definition}

@@

    Using our algorithm, the \ltl\ formula $\Phi = \always (a \limplies \eventually (\neg a))$ generates the un-simplified $\omega$-regular expression $((\lnot a) | (aa^*(\lnot a)))((\lnot a) | (aa^*(\lnot a)))^{\omega}$, and the simplified version $((\lnot a) | (aa^*(\lnot a)))^{\omega}$, which correspond to the two timelines in \cref{fig:unsimplified} and \cref{fig:simplified}.

@@

\begin{remark}
    Both simplified and un-simplified $\omega$-regular expressions would generate correct timeline representations. Correct as they faithfully represent the set of satisfying traces of the original \ltl\ formula $\Phi$. Nonetheless, we, as human users, oftentimes find the simplified version more intuitive to reason about.
\end{remark}

@@

\begin{figure}[h!]
    \centering
    \subfloat{{\includegraphics[width=6.5cm]{img/tllens-graph.png} }}%
    \quad % [\centering label 1]
    \subfloat{{\includegraphics[width=6.5cm]{img/star-height-graph.png}}}
    \caption{Visualization complexity metrics for formulas collected from Acacia~\cite{RV10} and NASA's  AAC~\cite{GCMTR16}}
    \label{fig:real-world}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=6.5cm]{img/binary.png}
    \caption{Scalability measured with binary counter examples from Rozier and Vardi~\cite{RV10}}
    \label{fig:scalability}
\end{figure}

@@

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from~\ref{ltl-defn} that we define \ltl\ formula as
\begin{align*}
    \Phi ::= & p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi \\ & \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi
\end{align*}
where $p \in \AP$. In our tool \tool, we use the following concrete syntax to represent \ltl\ formulae.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \Phi & ::=  & p                                     & \texttt{p}                                     & \text{atomic prop.} \\
         & \mid & \neg \Phi                             & \texttt{!} \Phi                                & \text{negation} \\
         & \mid & \Phi \land \Phi                       & \Phi \texttt{ \& } \Phi                        & \text{conjunction} \\
         & \mid & \Phi \lor \Phi                        & \Phi \texttt{ | } \Phi                         & \text{disjunction} \\
         & \mid & \Phi \limplies \Phi                   & \Phi \texttt{ -> } \Phi                        & \text{implication} \\
         & \mid & \always \Phi                          & \texttt{G} \Phi                                & \text{globally} \\
         & \mid & \eventually \Phi                      & \texttt{F} \Phi                                & \text{finally} \\
         & \mid & \nextt \Phi                           & \texttt{X} \Phi                                & \text{next} \\
         & \mid & \Phi \stronguntil \Phi                & \Phi \texttt{ U } \Phi                         & \text{until} \\
         & \mid & \Phi \weakrelease \Phi                & \Phi \texttt{ R } \Phi                         & \text{release} \\
    \end{array}
\]
%\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{RelatedWork}

>>>>

%\newcommand{\tool}{\hspace{0.1cm}\texttt{ltl2timeline}\hspace{0.1cm}}
\newcommand{\tool}{\hspace{0.1cm}\texttt{ltl2timeline}}

@@

\title{What's in a Name? \\ Linear Temporal Logic \\ Literally Represents Time Lines}

\author{\IEEEauthorblockN{Runming Li\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, PA, USA \\
runmingl@andrew.cmu.edu}
\and
\IEEEauthorblockN{Keerthana Gurushankar\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, PA, USA \\
kgurusha@andrew.cmu.edu}
\and
\IEEEauthorblockN{Marijn~J.H.~Heule}
\IEEEauthorblockA{\textit{Carnegie Mellon University}\\
Pittsburgh, PA, USA \\
marijn@cmu.edu}
\and
\IEEEauthorblockN{Kristin~Y.~Rozier}
\IEEEauthorblockA{\textit{Iowa State University} \\
Ames, IA, USA \\
kyrozier@iastate.edu}
}

@@

\thispagestyle{plain}
\pagestyle{plain}
\pagenumbering{gobble}

@@

\begin{definition}[Linear Temporal Logic (\ltl)]
    % cite Kristin's survey paper
    The syntax of an \ltl\ formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, is given by the following grammar:
    \begin{align*}
        \varphi ::= & p \mid \neg \varphi \mid \varphi \land \varphi \mid \varphi \lor \varphi \mid \varphi \limplies \varphi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \varphi \mid \varphi \weakrelease \varphi
    \end{align*}\label{ltl-defn}
\end{definition}

\begin{remark}
    Intuitively, $\always \varphi$ says that formula $\varphi$ is true at every time step; $\eventually \varphi$ says that formula $\varphi$ is true either now or at sometime in the future time step; and $\nextt \varphi$ says formula $\varphi$ is true at the next time step immediately after the current one; $\varphi \stronguntil \Psi$ says that formula $\varphi$ is true \textit{until} such time formula $\Psi$ becomes true; $\varphi \weakrelease \Psi$ says that formula $\Psi$ must be true now and remain true until formula $\varphi$ becomes true, after which point \textit{release} $\Psi$ (if formula $\varphi$ is never true, then formula $\Psi$ must remain true at all time).
\end{remark}

@@

\begin{definition}[Semantics of \ltl\ formula]
    The semantics of \ltl\ formula is defined via  $\pi, i \models \varphi$ where $\pi : \omega \rightarrow 2^{\AP}$ is the computation that stores the truthhood and falsehood of every atomic proposition at every time step, and $\omega$ is the set of natural numbers that denote the time step. The semantics follows exactly from Rozier~\cite{Roz11}.
\end{definition}

@@

    Using our algorithm, the \ltl\ formula $\varphi = \always (a \limplies \eventually (\neg a))$ generates the un-simplified $\omega$-regular expression $((\lnot a) | (aa^*(\lnot a)))((\lnot a) | (aa^*(\lnot a)))^{\omega}$, and the simplified version $((\lnot a) | (aa^*(\lnot a)))^{\omega}$, which correspond to the two timelines in \cref{fig:unsimplified} and \cref{fig:simplified}.

@@

\begin{remark}
    Both simplified and un-simplified $\omega$-regular expressions would generate correct timeline representations. Correct as they faithfully represent the set of satisfying traces of the original \ltl\ formula $\varphi$. Nonetheless, we, as human users, oftentimes find the simplified version more intuitive to reason about.
\end{remark}

@@

\begin{figure}[h!]
    \centering
    \subfloat{{\includegraphics[width=0.49\textwidth]{img/tllens-graph.png} }}%
    \quad % [\centering label 1]
    \subfloat{{\includegraphics[width=0.49\textwidth]{img/star-height-graph.png}}}
    \caption{Visualization complexity metrics for formulas collected from Acacia~\cite{RV10} and NASA's  AAC~\cite{GCMTR16}}
    \label{fig:real-world}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.49\textwidth]{img/binary.png}
    \caption{Scalability measured with binary counter examples from Rozier and Vardi~\cite{RV10}}
    \label{fig:scalability}
\end{figure}

@@

\section*{Acknowledgements}

This work was partially supported by the National Science Foundation (NSF) under grants CCF-2015445, CAREER-1664356, and CCRI-2016592.

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from~\ref{ltl-defn} that we define an \ltl\ formula as
\begin{align*}
    \varphi ::= & p \mid \neg \varphi \mid \varphi \land \varphi \mid \varphi \lor \varphi \mid \varphi \limplies \varphi \mid \always \varphi \mid \eventually \varphi \mid \nextt \varphi \\ & \mid \varphi \stronguntil \varphi \mid \varphi \weakrelease \varphi
\end{align*}
where $p \in \AP$. In our tool \tool, we use the following concrete syntax to represent \ltl\ formulae.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \varphi & ::=  & p                                     & \texttt{p}                                     & \text{atomic prop.} \\
         & \mid & \neg \varphi                             & \texttt{!} \varphi                                & \text{negation} \\
         & \mid & \varphi \land \varphi                       & \varphi \texttt{ \& } \varphi                        & \text{conjunction} \\
         & \mid & \varphi \lor \varphi                        & \varphi \texttt{ | } \varphi                         & \text{disjunction} \\
         & \mid & \varphi \limplies \varphi                   & \varphi \texttt{ -> } \varphi                        & \text{implication} \\
         & \mid & \always \varphi                          & \texttt{G} \varphi                                & \text{globally} \\
         & \mid & \eventually \varphi                      & \texttt{F} \varphi                                & \text{finally} \\
         & \mid & \nextt \varphi                           & \texttt{X} \varphi                                & \text{next} \\
         & \mid & \varphi \stronguntil \varphi                & \varphi \texttt{ U } \varphi                         & \text{until} \\
         & \mid & \varphi \weakrelease \varphi                & \varphi \texttt{ R } \varphi                         & \text{release} \\
    \end{array}
\]
%\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{RelatedWork}
